Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''

### 2019-10-16 writeup/pnas-word-order-si.tex
loglinear analysis

\paragraph{Other Greenberg Universals}
Greenberg~\cite{greenberg1963universals} stated a total of 45 universals.
Twenty of these concern the structure of individual words (as opposed to word order, which we focus on here), and many of those have been argued to be explained by the ``dual pressures'' idea \cite{haspelmath2006against}.
The other 25 universals concern word order; Dryer \cite{dryer1992greenbergian} reformulated most of these as correlations with verb-object order; these form the basis of our formalization in Table~\ref{table:greenberg-dryer}. 
There are a few other well-supported word order universals that are not correlations with verb-object order.
This includes dominance principles~\cite{croft2003typology} such as the strong preference for subjects to precede objects.
Furthermore, there has been interest in Greenberg's universals 18 and 20, which describe correlations not with verb-object order, but of different elements of noun phrases~\cite{cinque2005deriving, culbertson2014language, dryer2018order}.
Future work should examine whether these universals can also be linked to efficiency optimization.

@@

\paragraph{Comparison to other Formalizations}

\citet{justeson1990explanation} conducted a log-linear analysis on typological judgments of 147 languages, constructing an undirected graphical model.
%While they found evidence for correlations similar to those found by \citet{dryer1992greenbergian}, they found that some correlations are me
For the relations examined here, beyond correlations with verb-object order, they found additional correlations between (1) the noun-genitive and adposition-noun dependencies, and (2) between the noun-relative clause and adposition-noun dependencies, \emph{beyond} the correlation mediated through the individual correlations with the verb-object dependency.
%With their model, they found no evidence for a direct correlation V-O, G-N beyond that mediated through other correlations.
We ran the same Bayesian logistic mixed-effects analysis for these two correlations.
Results are shown in Figure~\ref{tab:corr-regression-implication}.
Both correlations are very strongly supported by grammars optimized for efficiency.
Future research might apply the graphical model approach of \citet{justeson1990explanation} to optimized grammars, and test whether grammar optimization predicts not just the presence of correlations, the structure their method found on typological data.

@@

@@

We asked whether efficiency optimization predicts dependency length minimization effects.
We first computed dependency length for grammars optimized for efficiency.
We found that 100\% of grammars optimized for efficiency reduce average dependency length compared to baseline grammars ($p < 0.05$, by one-sided $t$-test). 
This suggests that the reduction of dependency length observed in natural language is indeed predicted by efficiency maximization, confirming theoretical arguments made in prior work~\cite{rijkhoff1986word,rijkhoff1990explaining,hawkins1994performance,hawkins2004efficiency,futrell2017memory, futrell2017generalizing}.
Next, we constructed grammars that minimize average dependency length, using the same gradient descent method as we used for efficiency optimization (Section~\ref{sec:optim-eff}).
We expect that such grammars should have shorter dependency length than the real grammars, or grammars optimized for efficiency.
In Figure~\ref{fig:dlm-avg}, we plot the mean dependency length for optimized, real, and baseline orderings.
We find that optimizing grammars for efficiency reduces dependency length to a similar degree as found in the actual orderings in the corpora, almost up to the limit given by directly optimizing for dependency length. %[TODO acknowledge  Ferrer i Cancho R, Liu H (2014) The risks of mixing dependency lengths from se-quences of different length.Glottotheory5(2):143–155.]
We also plot more detailed results for four languages in Figure~\ref{fig:dlm-4langs}, plotting dependency length as a function of sentence length as reported in prior work \cite{futrell2015largescale}.
Optiziming grammars for efficiency produces dependency lengths similar to those found in the actual orderings.

@@

Next, we examined the word order properties of grammars optimized for DLM.
In Table~\ref{tab:all-predictions-1b}, we report the posterior prevalence of word order correlations in grammars optimized for DLM; our results show that optimizing for DLM makes predictions similar to efficiency optimization.
This is itself a novel result, suggesting that it is in part through favoring short dependencies that efficiency predicts word order universals, an idea that has been proposed in prior theoretical studies \citep{hawkins1994performance, hawkins2004efficiency}.
We find that these grammars also exhibit the eight correlations, similar to grammars directly optimized for efficiency.
On other correlations, predictions of DLM also resemble those of efficiency optimization.
However, it predicts strong correlations with \textit{amod} (adjectival modifiers) and \textit{nummod} (numeral modifiers) (see bottom of Table~\ref{tab:all-predictions-1b}), which are not borne out typologically.
In these cases, efficiency optimization predicts prevalences closer to 50\%, in line with typological data.

>>>>

\paragraph{Other Greenberg Universals}
Greenberg~\cite{greenberg1963universals} stated a total of 45 universals.
Twenty of these concern the structure of individual words (as opposed to word order, which we focus on here), and many of those have been argued to be explained by the ``dual pressures'' idea \cite{haspelmath2006against}.
The other 25 universals concern word order; Dryer \cite{dryer1992greenbergian} reformulated most of these as correlations with verb-object order; these form the basis of our formalization in Table~\ref{table:greenberg-dryer}. 
There are a few other well-supported word order universals that are not correlations with verb-object order.
This includes dominance principles~\cite{greenberg1963universals,croft2003typology} such as the strong preference for subjects to precede objects.
Furthermore, there has been interest in Greenberg's universals 18 and 20, which describe correlations not with verb-object order, but of different elements of noun phrases~\cite{cinque2005deriving, culbertson2014language, dryer2018order}.
Future work should examine whether these universals can also be linked to efficiency optimization.

@@

@@

\subsection{Comparison to other Formalizations of Greenberg's Correlations}

\citet{justeson1990explanation} conducted a log-linear analysis on typological judgments of 147 languages, constructing an undirected graphical model modeling correlations among any pair of six syntactic relations (verb-object, adposition-noun, noun-genitive, noun-relative clause, noun-adjective, verb-subject).
Results from their analysis suggested that, while some relations are directly correlated with the verb-object order, whereas other relations are only indirectly correlated with it.
In particular, in their analysis, the noun-genitive relation (corresponding to Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}} here) was not directly correlated with the verb-object correlation; instead, the typologically observed correlation was explained through correlations between the noun-genitive relation and other relations (such as the adposition-noun relation) that directly correlate with the verb-object relation.
Note that this does not contradict the statement that verb-object and noun-genitive relations correlate; it shows that the observed correlation can be explained through a chain of other correlations.

Since the set of syntactic relations examined here is different from that examined by \citet{justeson1990explanation}, we cannot directly compare the predictions of efficiency optimization with their results.
Nonetheless, we can show that efficiency optimization is compatible with a picture of Greenberg's correlation as a network of pairwise correlations among different syntactic relations, and in particular the result that the correlation between the verb-object and noun-genitive relations is mediated through other correlations.

First, we directly test the optimized grammars for two additional correlations found by \citet{justeson1990explanation}:
For the relations examined here, beyond correlations with verb-object order, they found additional correlations between (1) the noun-genitive and adposition-noun dependencies, and (2) between the noun-relative clause and adposition-noun dependencies, \emph{beyond} the correlation mediated through the individual correlations with the verb-object dependency.
%With their model, they found no evidence for a direct correlation V-O, G-N beyond that mediated through other correlations.
We ran the same Bayesian logistic mixed-effects analysis for these two correlations.
Results are shown in Figure~\ref{tab:corr-regression-implication}.
Both correlations are very strongly supported by grammars optimized for efficiency.

Second, we directly applied the log-linear analysis described by \citet{justeson1990explanation} to optimized grammars.
We represent each grammar via the directions $v_1, \dots v_9$ of the nine relations indicated in Table 1 of the main paper (verb-object, and \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {1}}}-\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}), we coded these as $-0.5$ for Japanese-like order, and $+0.5$ for Arabic-like order.
This analysis models the relative frequency $p_{(v_1, \dots, v_9)}$ of a particular configuration of such a configuration $(v_1, \dots, v_9)$ by a log-linear model:
\begin{equation}
	\log p_{(v_1, \dots, v_9)} = u_0 + \sum_{i=1}^9 u_i v_i + \sum_{i, j \in C} u_{i,j} v_{i} v_{j}
\end{equation}
where $C$ is some set of (unordered) pairs of relations $\in \{1, \dots, 9\}$, modeling those pairs of relations that directly correlate with each other, and where $u_0, i_i, i_{i,j}$ are real-valued parameters.
For instance, if all relations directly correlate with the verb-object order, and not with any other relation, $C$ would contain all the unordered pairs containing the verb-object relation.

We inferred the best-fitting such model by selecting the pairs in $C$ via forward-selection using AIC.
The best-fitting model includes a set $C$ of 13 correlating pairs, with $AIC=274.18$.
This resulting model is shown in Figure~\ref{fig:loglinear}; following \cite{justeson1990explanation}, we show those links between nodes that are included in this selected model.
In agreement with the results of \cite{justeson1990explanation}, a network is identified in which all relations are connected at least indirectly, but several relations are not directly connected to the verb-object relation:
In particular, in accordance with the typological data analysed by \cite{justeson1990explanation}, the observed correlation between the verb-object and noun-genitive relation is entirely mediated through correlations with other relations (adposition-noun and verb-adpositional phrase) that directly correlate with the verb-object relation.
A difference is that, in our analysis and unlike the analysis by \cite{justeson1990explanation}, the noun-relative clause dependency is not directly linked to the verb-object relation; this might be because our analysis takes a different set of relations into account compared to \cite{justeson1990explanation}.

We also note that, unlike our mixed-effects models, this log-linear model does not have random effects, as we found that adding random effects to the log-linear model led to nonconvergence.
This means that it does not account for differences in the tree structures between languages and language families; as a result, the mixed-effects analyses for individual correlation pairs may be more conservative than this log-linear model.
Future work should replicate the analysis of \cite{justeson1990explanation} on a larger typological database and with more relations, to enable a direct comparison with the network structure predicted by  efficiency optimization.

\begin{figure}
	\begin{center}
	\includegraphics[width=0.8\textwidth]{../results/correlations/analysis/controls/loglinear-pairwise-correlations.pdf}
	\end{center}
	\caption{Network of pairwise correlations among the nine syntactic relations examined in this study, estimated from grammars optimized, identified using a log-linear model following \citet{justeson1990explanation}. The verb-object relation is at the center of the network. Relations between verbs and their dependents are colored in blue; relations between nouns and their dependents are colored in red; other relations are colored in green.}\label{fig:loglinear}
\end{figure}

@@

We asked whether efficiency optimization predicts dependency length minimization effects.
We first computed dependency length for grammars optimized for efficiency.
We found that 100\% of grammars optimized for efficiency reduce average dependency length compared to baseline grammars ($p < 0.05$, by one-sided $t$-test). 
This suggests that the reduction of dependency length observed in natural language is indeed predicted by efficiency maximization, confirming theoretical arguments made in prior work~\cite{hawkins1994performance,hawkins2004efficiency,futrell2017memory, futrell2017generalizing}.
Next, we constructed grammars that minimize average dependency length, using the same gradient descent method as we used for efficiency optimization (Section~\ref{sec:optim-eff}).
We expect that such grammars should have shorter dependency length than the real grammars, or grammars optimized for efficiency.
In Figure~\ref{fig:dlm-avg}, we plot the mean dependency length for optimized, real, and baseline orderings.
We find that optimizing grammars for efficiency reduces dependency length to a similar degree as found in the actual orderings in the corpora, almost up to the limit given by directly optimizing for dependency length. %[TODO acknowledge  Ferrer i Cancho R, Liu H (2014) The risks of mixing dependency lengths from se-quences of different length.Glottotheory5(2):143–155.]
We also plot more detailed results for four languages in Figure~\ref{fig:dlm-4langs}, plotting dependency length as a function of sentence length as reported in prior work \cite{futrell2015largescale}.
Optiziming grammars for efficiency produces dependency lengths similar to those found in the actual orderings.

@@

Next, we examined the word order properties of grammars optimized for DLM.
In Table~\ref{tab:all-predictions-1b}, we report the posterior prevalence of word order correlations in grammars optimized for DLM; our results show that optimizing for DLM makes predictions similar to efficiency optimization.
We find that these grammars also exhibit the eight correlations, similar to grammars directly optimized for efficiency.
This is itself a novel result, suggesting that it is in part through favoring short dependencies that efficiency predicts word order universals, an idea that has been proposed in prior theoretical studies, though never tested computationally on large-scale text data \citep{kuno1974position,rijkhoff1986word,frazier1985syntactic,rijkhoff1990explaining,hawkins1994performance, hawkins2004efficiency}.
On other correlations, predictions of DLM also resemble those of efficiency optimization.
However, it predicts strong correlations with \textit{amod} (adjectival modifiers) and \textit{nummod} (numeral modifiers) (see bottom of Table~\ref{tab:all-predictions-1b}), which are not borne out typologically.
In these cases, efficiency optimization predicts prevalences closer to 50\%, in line with typological data.
